---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
Find data file in local branch of repo, unzip file and read as 'dat'

```{r}
        datfile <- unzip("~/GitHub/RepData1/RepData_PeerAssessment1/activity.zip")
        dat <- read.csv("./activity.csv", stringsAsFactors = FALSE)
```

View file and make summary

```{r}
        head(dat)
        summary(dat)
```
Prepare data for analysis - create 'day-of-the-week' variable 'dotw' 

```{r}
library(lubridate)
        dat$dotw <- ymd(dat$date)
        dat$dotw <- wday(dat$dotw)
        head(dat)
```


## What is mean total number of steps taken per day?

Find number of days in data set

```{r}
        days <- as.data.frame(unique(dat$date))
        nrow(days)
```
To avoid duplicate days of the month, create unique day counter variable from date

```{r}
         dat$span <- interval(ymd("2012-10-01"), ymd(dat$date))
         dat$daycount <- as.numeric(dat$span,"days") + 1
         head(dat)
```

group dat by daycount and find total steps per day and plot histogram

```{r}
        library(dplyr)
        fulldays <- group_by(dat, daycount)
        totals <- summarise(fulldays, totsteps=sum(steps, na.rm = TRUE))
        head(totals)
        hist(totals$totsteps, main="Total Steps Per Day",xlab="Total Steps",ylab="Frequency", plot = TRUE)
```

find mean and median value of 'totals'

```{r}
        meantotsteps <- mean(totals$totsteps)
        # mean total number of steps taken per day :
        meantotsteps
        mediantotsteps <- median(totals$totsteps)
        # median total number of steps taken per day:
        mediantotsteps
                                
```

## What is the average daily activity pattern?

Group by interval and calculate mean value for each interval.  Convert intervals to time for plotting. Plot mean v/s time of day to look for patterns.

```{r}
        inter <- group_by(dat, interval)  # group by interval
        AvIntSteps <- summarise(inter, avSteps = mean(steps, na.rm = TRUE))  # calculate mean of each group
        head(AvIntSteps)  # check data
        ais <- AvIntSteps  # create shorter df name for easy typing!
        ais$id <- seq.int(nrow(ais))  # create new column using index number
        ais$int2 <- ais$id*5-5  # create new column showing minutes since mid-night
        ais$hour <- ais$int2/60  # convert minutes to hours and decimals
        head(ais)   # check data again
        plot(ais$hour,ais$avSteps, main= "Activity Trend", xlab="time (hrs - 24hr clock)", ylab="Average Steps per 5 minute Increment")  # make simple scatter plot to detect trends
        maxsteps <- max(ais$avSteps)  # find maximum value for average steps
        maxint <- filter(ais, avSteps==maxsteps)  # find row where this occurs
        maxint$interval    # display interval
```

The average daily activity pattern shows minimal activity during the late evening and night time hours, fairly constant activity throughout the day - with a pronounced spike from about 8:00 a.m.to 10:00 a.m.
        
## Imputing missing values

Find number of NAs in dat$steps and then check data frame to see if there are more in other columns

```{r}

        nodat <- sum(is.na(dat$steps))
        # number of NAs in steps column is :
        nodat
        nodat2 <- sum(is.na(dat))
        # total NAs in dat is :
        nodat2
```
All the NAs are in the 'steps' column

Substitute interval means for NAs.  Merge dfs 'dat' and 'ais' using interval and create new column of interval means 'intmean'.  Then create new steps2 column with NAs set to 'intmean'.  Analyse new data set.

```{r}
        dat$intmean <- ais$avSteps[match(dat$interval,ais$interval)] 
        dmat <- as.matrix(dat)  #convert df to matrix
        y <- which(is.na(dmat)==TRUE)  # identify all the NA rows
        dmat[y] <- dmat[y,7]
        head(dmat)
        dat2 <- as.data.frame(dmat)
             
        fulldays2 <- group_by(dat2, daycount)
        totals2 <- summarise(fulldays2, totsteps2=sum(as.numeric(steps)))
        head(totals2)
        hist(totals2$totsteps2, main="Total Steps Per Day (NA substituted)",xlab="Total Steps",ylab="Frequency", plot = TRUE)
```


Calculate the new mean and median of total steps per day

```{r}

        meantotsteps2 <- mean(as.numeric(totals2$totsteps2))
        # mean total number of steps taken per day :
        meantotsteps2
        mediantotsteps2 <- median(as.numeric(totals2$totsteps2))
        # median total number of steps taken per day:
        mediantotsteps2

```









## Are there differences in activity patterns between weekdays and weekends?
