---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
Find data file in local branch of repo, unzip file and read as 'dat'

```{r}
        datfile <- unzip("~/GitHub/RepData1/RepData_PeerAssessment1/activity.zip")
        dat <- read.csv("./activity.csv", stringsAsFactors = FALSE)
```

View file and make summary

```{r}
        head(dat)
        summary(dat)
```
Prepare data for analysis - create 'day-of-the-week' variable 'dotw' 

```{r}
library(lubridate)
        dat$dotw <- ymd(dat$date)
        dat$dotw <- wday(dat$dotw)
        head(dat)
```


## What is mean total number of steps taken per day?

Find number of days in data set

```{r}
        days <- as.data.frame(unique(dat$date))
        nrow(days)
```
To avoid duplicate days of the month, create unique day counter variable from date

```{r}
         dat$span <- interval(ymd("2012-10-01"), ymd(dat$date))
         dat$daycount <- as.numeric(dat$span,"days") + 1
         head(dat)
```

group dat by daycount and find total steps per day and plot histogram

```{r}
        library(dplyr)
        fulldays <- group_by(dat, daycount)
        totals <- summarise(fulldays, totsteps=sum(steps, na.rm = TRUE))
        head(totals)
        hist(totals$totsteps, main="Total Steps Per Day",xlab="Total Steps",ylab="Frequency", plot = TRUE)
```

find mean and median value of 'totals'

```{r}
        meantotsteps <- mean(totals$totsteps)
        # mean total number of steps taken per day :
        meantotsteps
        mediantotsteps <- median(totals$totsteps)
        # median total number of steps taken per day:
        mediantotsteps
                                
```

## What is the average daily activity pattern?

Group by interval and calculate mean value for each interval.  Convert intervals to time for plotting. Plot mean v/s time of day to look for patterns.

```{r}
        inter <- group_by(dat, interval)  # group by interval
        AvIntSteps <- summarise(inter, avSteps = mean(steps, na.rm = TRUE))  # calculate mean of each group
        head(AvIntSteps)  # check data
        ais <- AvIntSteps  # create shorter df name for easy typing!
        ais$id <- seq.int(nrow(ais))  # create new column using index number
        ais$int2 <- ais$id*5-5  # create new column showing minutes since mid-night
        ais$hour <- ais$int2/60  # convert minutes to hours and decimals
        head(ais)   # check data again
        plot(ais$hour,ais$avSteps, main= "Activity Trend", xlab="time (hrs - 24hr clock)", ylab="Average Steps per 5 minute Increment")  # make simple scatter plot to detect trends
        maxsteps <- max(ais$avSteps)  # find maximum value for average steps
        maxint <- filter(ais, avSteps==maxsteps)  # find row where this occurs
        maxint$interval    # display interval
```

The average daily activity pattern shows minimal activity during the late evening and night time hours, fairly constant activity throughout the day - with a pronounced spike from about 8:00 a.m.to 10:00 a.m.
        
## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?
